Type theory in type theory with single substitutions

### 2024-09-18 types_2024_postproc/p.tex
stuff

\begin{abstract}
Type theory can be described as a generalised algebraic theory. This
automatically gives a notion of model and the existence of the syntax
as the initial model, which is a quotient inductive-inductive
type. Algebraic definitions of type theory include Ehrhard's
definition of model, categories with families (CwFs), contextual
categories, Awodey's natural models, C-systems, B-systems. With the
exception of B-systems, these notions are based on a parallel
substitution calculus where substitutions form a category. In this
paper we define a single substitution calculus (SSC) for type theory
and show that the SSC syntax and the CwF syntax are isomorphic for a
theory with dependent function space and a hierarchy of universes. SSC
only includes single substitutions and single weakenings, and eight
equations relating these: four equations describe how to substitute
variables, and there are four equations on types which are needed to
typecheck other equations. All the other equations are substitution
(naturality) rules or computation rules for different type
formers. SSC provides a simple, minimalistic alternative to parallel
substitution calculi or B-systems for defining type theory. SSC
relates to CwF as extensional combinatory calculus relates to lambda
calculus. In case of additional type formers, we show that SSC models
are actually equivalent to CwF-based models in a weak sense. Most
results in this paper were formalised in Agda.
\end{abstract}

@@

Type theory is a language which can be described as a second-order
generalised algebraic theory (SOGAT
\cite{DBLP:conf/fscd/KaposiX24}). Some properties of this description:
\begin{enumerate}
\item[(i)] It is an intrinsic presentation
  \cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16} as
  opposed to extrinsic \cite{abel2013normalization,theo}. That is, we
  only consider well-formed, well-scoped and well-typed abstract
  syntax trees, there are no meaningless terms.
\item[(ii)] Terms are quotiented by the conversion relation.
\item[(iii)] Every operation is stable under substitution.
\end{enumerate}
For example, type theory with $\Pi$ types and universes {\`a} la
Coquand \cite{coquandUniverse} is specified by the following SOGAT.
\begin{equation}\label{eq:tt}
\begin{alignedat}{10}
  & \Ty && : \N\ra\Set \\
  & \Tm && : \Ty\,i\ra\Set \\
  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
  & \U && : (i:\N)\ra\Ty\,(1+i) \\
  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd \\
  & \Lift && : \Ty\,i\ra\Ty\,(1+i) \\
  & \un && : \Tm\,(\Lift\,A)\cong\Tm\,A : \mk
\end{alignedat}
\end{equation}
Because types are isomorphic to a term (witnessed by $\El$ and $\cd$),
we have to index types by their (universe) level, and types at lower
levels are included at upper levels by $\Lift$. We use implicit
arguments, e.g.\ $\Tm$ takes the $i:\N$ implicitly, and $\app$ takes
$i$, $A$ and $B$ implicitly. $f : X \cong Y : g$ denotes an
isomorphism, that is, $f : X\ra Y$, $g:Y\ra X$ with $g\,(f\,x) = x$
and $f\,(g\,y) = y$ for all $x$, $y$. Note that $\Pi$ and $\lam$ are
second-order functions. We say that $\Pi$ binds a $\Tm$-variable of
type $A$ in its second $\Ty$-argument. Every binder is represented by
a second-order function, and as a result, we don't need to talk about
contexts and substitutions when describing the theory: we re-use the
binding features of the metatheory to describe those of our object
theory (this technique is also known as higher-order abstract syntax
\cite{DBLP:conf/lics/Hofmann99} or logical framework
\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2} and
our notation comes from two-level type theory \cite{DBLP:conf/csl/AltenkirchCK16,DBLP:journals/mscs/AnnenkovCKS23}). There is no good notion of homomorphism between
second-order models, thus the semantics of a SOGAT is given by a first
order generalised algebraic theory (GAT
\cite{DBLP:journals/apal/Cartmell86}). GATs are generalisations of
single-sorted algebraic theories (such as monoids, groups) to multiple
sorts where later sorts can be indexed over previous sorts. The usual
results from universal algebra (e.g.\ models form a category with
finite limits and initial object, existence of free and co-free
models) carry over to GATs \cite{andras,DBLP:phd/hal/Moeneclaey22}. An
example of a GAT is the theory of categories where there is a sort of
objects and a sort of morphisms double-indexed over objects.

@@

In the parallel semantics, the GAT always starts with a category with
a terminal object, objects of the category are called contexts,
morphisms are substitutions. Every sort/operation of the SOGAT gives
rise to a sort/operation in the GAT which is now indexed by
contexts. The sorts moreoever come with instantiation (sometimes
called substitution) operations which are functorial, and all the
operations are natural with respect to instantiation. For those sorts
which are bound by an operation in the SOGAT description, we have
context extension, meaning that contexts can carry variables of that
sort (in other words, the sort is given by a locally representable
presheaf). The parallel semantics of the SOGAT
\begin{equation}\label{eq:tytm}
\begin{alignedat}{10}
  & \Ty && : \Set \\
  & \Tm && : \Ty \ra\Set 
\end{alignedat}
\end{equation}
is the GAT of categories with families (CwFs
\cite{DBLP:conf/types/Dybjer95,Castellan2021}, Definition \ref{def:cwf}).
We define the notion ``model of the SOGAT'' by its parallel substitution
semantics, which is a GAT, and that has an obvious notion of model. The
syntax of the language is the initial model, which exists for every
GAT \cite{DBLP:journals/pacmpl/KaposiKA19}. This is an intrinsic
syntax (only well-typed terms), quotiented by the equations in the
SOGAT, which describe conversion for the language.

The single substitution semantics is a GAT which does not start with a
category: there are contexts and maps between contexts, but these maps
are single substitutions or single weakenings, they don't
compose. There are fewer operations and fewer equations than in the
parallel semantics. In this paper we investigate the single
substitution semantics of the type theory specified by the SOGAT
(\ref{eq:tt}). This presentation of type theory is simpler than CwFs,
the operations are all easy to motivate, we illustrate this in Section
\ref{sec:tt}. Single substitution calculi are closer to the usual
presentation of type theory using derivation rules, this was one of
the motivations of Voevodsky for developing B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}. Our single substitution
semantics of (\ref{eq:tytm}) is simpler and has more models than
B-systems. GAT presentations of type theory are important for defining
models. Our presentation gives a new notion of model, which in the
future might contribute to definitions of new models. It can also
influence implementations of type theory. Because our single
substitution GAT does not involve a category, it might be easier to
adapt it to a higher dimensional setting, see Section
\ref{sec:conclusion}.

>>>>

\begin{abstract}
Type theory can be described as a generalised algebraic theory. This
automatically gives a notion of model and the existence of the syntax
as the initial model, which is a quotient inductive-inductive
type. Algebraic definitions of type theory include Ehrhard's
definition of model, categories with families (CwFs), contextual
categories, Awodey's natural models, C-systems, B-systems. With the
exception of B-systems, these notions are based on a parallel
substitution calculus where substitutions form a category. In this
paper we define a single substitution calculus (SSC) for type theory
and show that the SSC syntax and the CwF syntax are isomorphic for a
theory with dependent function space and a hierarchy of universes. SSC
only includes single substitutions and single weakenings, and eight
equations relating these: four equations describe how to substitute
variables, and there are four equations on types which are needed to
typecheck other equations. All the other equations are substitution
(naturality) rules or computation rules for different type
formers. SSC provides a simple, minimalistic alternative to parallel
substitution calculi or B-systems for defining type theory. SSC
relates to CwF as extensional combinatory calculus relates to lambda
calculus. If we have some additional type formers, we show that SSC models
are actually equivalent to CwF-based models in a weak sense. Most
results in this paper were formalised in Agda.
\end{abstract}

@@

Type theory is a language which can be described as a second-order
generalised algebraic theory (SOGAT
\cite{DBLP:conf/fscd/KaposiX24}). Some properties of this description:
\begin{enumerate}
\item[(i)] It is an intrinsic presentation
  \cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16} as
  opposed to extrinsic \cite{abel2013normalization,theo}. That is, we
  only consider well-formed, well-scoped and well-typed abstract
  syntax trees, there are no meaningless terms.
\item[(ii)] Terms are quotiented by the conversion relation.
\item[(iii)] Every operation is stable under substitution.
\end{enumerate}
For example, type theory with $\Pi$ types and universes {\`a} la
Coquand \cite{coquandUniverse} is specified by the following SOGAT.
\begin{equation}\label{eq:tt}
\begin{alignedat}{10}
  & \Ty && : \N\ra\Set \\
  & \Tm && : \Ty\,i\ra\Set \\
  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
  & \U && : (i:\N)\ra\Ty\,(1+i) \\
  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd \\
  & \Lift && : \Ty\,i\ra\Ty\,(1+i) \\
  & \un && : \Tm\,(\Lift\,A)\cong\Tm\,A : \mk
\end{alignedat}
\end{equation}
Because types are isomorphic to certain terms (witnessed by $\El$ and $\cd$),
we have to index types by their (universe) level, and types at lower
levels are included at upper levels by $\Lift$. We use implicit
arguments, e.g.\ $\Tm$ takes the $i:\N$ implicitly, and $\app$ takes
$i$, $A$ and $B$ implicitly. $f : X \cong Y : g$ denotes an
isomorphism, that is, $f : X\ra Y$, $g:Y\ra X$ with $g\,(f\,x) = x$
and $f\,(g\,y) = y$ for all $x$, $y$. Note that $\Pi$ and $\lam$ are
second-order functions. We say that $\Pi$ binds a $\Tm$-variable
in its second $\Ty$-argument. Every binder is represented by
a second-order function, and as a result, we don't need to talk about
contexts and substitutions when describing the theory: we re-use the
binding features of the metatheory to describe those of our object
theory (this technique is also known as higher-order abstract syntax
\cite{DBLP:conf/lics/Hofmann99}, logical framework
\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2} or
two-level type theory \cite{DBLP:conf/csl/AltenkirchCK16,DBLP:journals/mscs/AnnenkovCKS23}). There is no good notion of homomorphism between
second-order models, thus the semantics of a SOGAT is given by a first
order generalised algebraic theory (GAT
\cite{DBLP:journals/apal/Cartmell86}). GATs are generalisations of
single-sorted algebraic theories (such as monoids, groups) to multiple
sorts where later sorts can be indexed over previous sorts. The usual
results from universal algebra (e.g.\ models form a category with
finite limits and initial object, existence of free and co-free
models) carry over to GATs \cite{andras,DBLP:phd/hal/Moeneclaey22}. An
example of a GAT is the theory of categories where there is a sort of
objects and a sort of morphisms double-indexed over objects.

@@

In the parallel semantics, the GAT always starts with a category with
a terminal object, objects of the category are called contexts,
morphisms are substitutions. Every sort/operation of the SOGAT gives
rise to a sort/operation in the GAT which is now indexed by
contexts. The sorts moreover come with instantiation (sometimes
called substitution) operations which are functorial, and all the
operations are natural with respect to instantiation. For those sorts
which are bound by an operation in the SOGAT description, we have
context extension, meaning that contexts can carry variables of that
sort (in other words, the sort is given by a locally representable
presheaf). The parallel semantics of the SOGAT
\begin{equation}\label{eq:tytm}
\begin{alignedat}{10}
  & \Ty && : \Set \\
  & \Tm && : \Ty \ra\Set 
\end{alignedat}
\end{equation}
is the GAT of categories with families (CwFs
\cite{DBLP:conf/types/Dybjer95,Castellan2021}, Definition \ref{def:cwf}).
We define the notion ``model of the SOGAT'' by its parallel substitution
semantics, which is a GAT, and that has an obvious notion of model. The
syntax of the language is the initial model, which exists for every
GAT \cite{DBLP:journals/pacmpl/KaposiKA19}. This is an intrinsic
syntax (only well-typed terms), quotiented by the equations in the
SOGAT, which describe conversion for the language.

The single substitution semantics is a GAT which does not start with a
category: there are contexts and maps between contexts, but these maps
are single substitutions or single weakenings, they don't
compose. There are fewer operations and fewer equations than in the
parallel semantics. In this paper we investigate the single
substitution semantics of the type theory specified by the SOGAT
(\ref{eq:tt}). This presentation of type theory is simpler than CwFs,
the operations are all easy to motivate, we illustrate this in Section
\ref{sec:tt}. Single substitution calculi are closer to the usual
presentation of type theory using derivation rules, this was one of
the motivations of Voevodsky for developing B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}. Our single substitution
semantics of (\ref{eq:tytm}) is simpler and has more models than
B-systems.
